Cancer Applications of Choropleth Maps, and the Potential of Cartograms and Alternative Map Displays

### 2019-12-17 review_article.Rmd
updated plots

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

<!-- Measures reported in cancer atlases -->
Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the public and others not trained in statistical analysis.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analysis. Cruickshank (1947) as cited by S. D. Walter [@DMAHP], discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps used to communicate to members of the public and other non-expert audiences are choropleth maps.

Epidemiologists and statisticians have developed the statistics used to communicate the burden of cancer over several decades. Table 1<!--\ref{tab:measures}--> summarizes the measures commonly presented in published cancer atlases. Mortality rates are commonly presented as relative rates of risk across the population and age-adjusted to correct for the higher prevalence of cancers in older populations.
Howe [@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s.

Table: 1<!--\label{tab:measures}--> Common measures for reporting cancer information.

@@

<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 cancer atlases published between 2010 and 2018. Each of these online atlases uses choropleth maps. All except one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships. Figure 1<!--\ref{fig:choropleth-grid-create}--> displays a subset of maps from these atlases, the selection varies in the geographies explored. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows Globocan 2012 [@Globocan] which explores Estimated Cancer Incidence, Mortality and Prevalence Worldwide using data sourced from cancer registries of each country. 
The Bowel Cancer Australia Atlas in Figure 1<!--\ref{fig:choropleth-grid-create}-->d presents an example of a cancer specific atlas, it shows the average Standardized Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 in Australia [@Bowel]. Like many of the atlases examined, there is a choice of gender displayed in the Bowel Cancer Atlas. An alternative approach used in the Map of Cancer Mortality Rates in Spain, Figure 1<!--\ref{fig:choropleth-grid-create}-->f shows side-by-side maps of the relative risk of lung cancer separately for men and women [@cancerSpain].

There are vast differences in the geographies considered by online atlases. 
Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows global age-standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018.
The United States Cancer Statistics [@USInteractive] presented in Figure 1<!--\ref{fig:choropleth-grid-create}-->e contains the incidence rate per 100,000, of all cancer types for men and women in the United States in 2016, aggregated at the state level.
The Environment and Health Atlas of England and Wales, Figure 1<!--\ref{fig:choropleth-grid-create}-->a shows the relative risk for women developing lung cancer in England and Wales in 2010 [@EnvEnglandWales2] at a neighborhood (small-area) scale.
The Atlas of Cancer in Queensland, Figure 1<!--\ref{fig:choropleth-grid-create}-->c shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 in the state of Queensland within Australia [@QLDcancerAtlas].

@@

Australia presents an extreme case of an urban rural divide. The land mass occupied by urban electoral districts is only 10% of Australia, yet 90% of the population live in these urban areas [@ACTUC]. This impedes reading information about disease incidence from choropleth maps.

@@

@@

Table: 2<!--\label{tab:usa}--> Maps used to present statistics for the United States of America. The colour of each state communicates the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016.

@@

Figure 2<!--\ref{fig:ggcartograms}--> shows four different cartograms for the same data. The information in Table 2<!--\ref{tab:usa}--> summarizes what can be observed in the four types of cartograms.

@@

Figure 3<!--\ref{fig:auscartograms}--> shows four main types of cartograms displaying melanoma incidence on Statistical Areas at Level 3 [@ABSmaps] areas in Australia. The version of a contiguous cartogram (a) has expanded the highly populated areas while preserving the full shapes of rural areas. It has not fully resolved the population transformation of areas, and if it had accurately sized areas by population, the country would be unrecognizable. The shape-preserved cartogram is unreadable, and it has reduced all areas to tiny spots on the map. Zooming in on a high-resolution output shows it does preserve the areas. The Dorling cartogram has lost all geographic context. The hexagon tile map provides a reasonable spatial distribution despite having too much white-space in the outback areas.

@@

Some mix of tiling, faceting or even micro maps, which allow some spatial continuity while also zooming into small areas, are good solutions for difficult geographies. Table \ref{tab:methods} summarizes the key criteria for testing maps and alternative displays. Moore and Carpenter [@SAMGIS] and Bell et al. [@CPISACA] provide suggestions and comments to help map creators best communicate their health data and spatial analysis.

@@

Table: \label{tab:methods} Summary of features and constraints of common mapping methods used to display cancer statistics (Y=Yes, N=No, S=Sometimes).

>>>>

<!-- Measures reported in cancer atlases -->
Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the public and others not trained in statistical analysis.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analysis. Cruickshank (1947) as cited by S. D. Walter [@DMAHP], discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, choropleth maps are visulisation chosen to communicate cancer statistics to members of the public and other non-expert audiences.

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends. These atllases are described in Table 1.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

Table: 1<!--\label{tab:choropleths}--> A selection of choropleth cancer maps from online atlases.

| Fig. | Atlas | Statistic | Data source | 
|:----:|:-------------|:----------------------|:----------------|  
|1a | The Environment and Health Atlas of England and Wales | relative risk for women developing lung cancer in England and Wales in 2010 [@EnvEnglandWales2]| [ Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU) ](http://www.envhealthatlas.co.uk/homepage/)|
|1b | Globocan 2018: Estimated Cancer Incidence, Mortality and Prevalence Worldwide | age standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018 [@Globocan]| [World Health Organization's International Agency for Research on Cancer](https://www.iarc.fr/). |
|1c | Atlas of Cancer in Queensland | the relative incidence ratio of lung cancer in males in the state of QLD within Australia based on data from 1998 to 2007, Queensland Cancer Council[@QLDcancerAtlas] | [Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/) |
|1d | Bowel Cancer Australia Atlas | the percentage of Australian males between 50 - 54 years of age diagnosed with bowel cancer in 2016 in Australia [@Bowel]. |  [Bowel Cancer Australia](https://www.bowelcanceraustralia.org/)|
|1e | United States Cancer Statistics: An Interactive Cancer Statistics Website | the incidence rate per 100,000, of all cancer types for men and women in the United States in 2016, aggregated at the state level [@USInteractive].|  *[Centers for Disease Control and Prevention](https://www.cdc.gov/)*, with data from state cancer registries. | 
|1f | Map of Cancer Mortality Rates in Spain | side by side maps of relative risk of lung cancer for men vs women for 2004 to 2008 [@cancerSpain].| [Map of cancer mortality rates in Spain](https://elpais.com/elpais/2014/10/06/media/1412612722_141933.html)|
|1g | Atlas of Childhood Cancer in Ontario | the incidence rate of childhood cancers per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004 [@OntarioPediatric]. | [The Pediatric Oncology Group of Ontario Networked Information System](https://www.pogo.ca/research-data/pogonis-childhood-cancer-database/data-requests/)|

Epidemiologists and statisticians have developed the statistics used to communicate the burden of cancer over several decades. Table 2<!--\ref{tab:measures}--> summarizes the measures commonly presented in published cancer atlases. Mortality rates are commonly presented as relative rates of risk across the population and age-adjusted to correct for the higher prevalence of cancers in older populations.
Howe [@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s.

Table: 2<!--\label{tab:measures}--> Common measures for reporting cancer information.

@@

<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 cancer atlases published between 2010 and 2018. Each of these online atlases uses choropleth maps. All except one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships. Figure 1<!--\ref{fig:choropleth-grid-create}--> displays a subset of maps from these atlases, the selection varies in the geographies explored. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows Globocan 2018 [@Globocan] which explores Estimated Cancer Incidence, Mortality and Prevalence Worldwide using data sourced from cancer registries of each country. 
The Bowel Cancer Australia Atlas in Figure 1<!--\ref{fig:choropleth-grid-create}-->d presents an example of a cancer specific atlas, it shows the average Standardized Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 in Australia [@Bowel]. Like many of the atlases examined, there is a choice of gender displayed in the Bowel Cancer Atlas. An alternative approach used in the Map of Cancer Mortality Rates in Spain, Figure 1<!--\ref{fig:choropleth-grid-create}-->f shows side-by-side maps of the relative risk of lung cancer separately for men and women [@cancerSpain].

There are vast differences in the geographies considered by online atlases. 
Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows global age-standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018.
The United States Cancer Statistics [@USInteractive] presented in Figure 1<!--\ref{fig:choropleth-grid-create}-->e contains the incidence rate per 100,000, of all cancer types for men and women in the United States in 2016, aggregated at the state level.
The Environment and Health Atlas of England and Wales, Figure 1<!--\ref{fig:choropleth-grid-create}-->a shows the relative risk for women developing lung cancer in England and Wales in 2010 [@EnvEnglandWales2] at a neighborhood (small-area) scale.
The Atlas of Cancer in Queensland, Figure 1<!--\ref{fig:choropleth-grid-create}-->c shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 [@abs2016] in the state of Queensland within Australia [@QLDcancerAtlas].

@@

Australia presents an extreme case of an urban rural divide. The land mass occupied by urban electoral districts is only 10% of Australia, yet 90% of the population live in these urban areas [@ACTUC].

@@

Choropleth maps colour each geographic unit to allow map users to measure the value of the statistic [@EI]. Map users contrast the colours in neighbouring areas to understand the spatial distribution. Map creators must choose an attractive and effective colour scheme, but is a difficult task [@CB]. The ColorBrewer system supports map creators in the decision making process, but this does not consider perceptual colour
specifications.

Brewer [@CB] also suggests the use and colors of borders and backgrounds can change the appearance of the colors representing the value of the statistics.
Incorporating borders may allow a reference point to orient users as they compare colours of neighbouring areas.

@@

Table: 3<!--\label{tab:usa}--> Maps used to present statistics for the United States of America. The colour of each state communicates the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016.

@@

Figure 2<!--\ref{fig:ggcartograms}--> shows four different cartograms for the same data. The information in Table 3<!--\ref{tab:usa}--> summarizes what can be observed in the four types of cartograms.

@@

Figure 3<!--\ref{fig:auscartograms}--> shows four main types of cartograms displaying melanoma incidence on Statistical Areas at Level 3 [@abs2016] in Australia. The version of a contiguous cartogram (a) has expanded the highly populated areas while preserving the full shapes of rural areas. It has not fully resolved the population transformation of areas, and if it had accurately sized areas by population, the country would be unrecognizable. The shape-preserved cartogram is unreadable, and it has reduced all areas to tiny spots on the map. Zooming in on a high-resolution output shows it does preserve the areas. The Dorling cartogram has lost all geographic context. The hexagon tile map provides a reasonable spatial distribution despite having too much white-space in the outback areas.

@@

Some mix of tiling, faceting or even micro maps, which allow some spatial continuity while also zooming into small areas, are good solutions for difficult geographies. Table 3 <!--\ref{tab:methods}--> summarizes the key criteria for testing maps and alternative displays. Moore and Carpenter [@SAMGIS] and Bell et al. [@CPISACA] provide suggestions and comments to help map creators best communicate their health data and spatial analysis.

@@

Table: 3<!--\label{tab:methods}--> Summary of features and constraints of common mapping methods used to display cancer statistics (Y=Yes, N=No, S=Sometimes).
